Geometric models reveal behavioral and neural signatures of how naturalistic experiences are transformed into episodic memories

### 2020-08-26 paper/main.tex
inputted JRM notes to results section

Occasionally, this gradual ``drift" of our ongoing experience is punctuated by sudden changes, or ``shifts"~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) help to discretize our experiences (and their mental representations) into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within-event) and transient (across-event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distills those experiences down to their essence.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.  This work also suggests a means of distinguishing the essence of an experience from its low-level details.  The overall structure of events and event transitions reflects how the high-level experience unfolds (i.e., its essence), while subtler event-level properties reflect low-level details.  Prior research has also implicated a network of brain regions (including the hippocampus and the medial prefrontal cortex) in playing a critical role in transforming experiences into structured and consolidated memories ~\citep{TompDava17}.

@@

\cite{ChenEtal17} collected hand-annotated information about each of 1,000 (manually identified) scenes spanning the roughly 50 minute video used in their experiment.  This information included: a brief narrative description of what was happening, the location where the scene took place, the names of any characters on the screen, and other similar details (for a full list of annotated features, see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features and scenes as the ``vocabulary'' for the topic model.  We then concatenated the sets of words across all features contained in overlapping sliding windows of (up to) 50 scenes, and treated each window as a single ``document" for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the episode (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  Note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06}, in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each sliding window of annotations transformed by the topic model.  We then stretched (interpolated) the resulting windows-by-topics matrix to match the time series of the 1,976 fMRI volumes collected as participants viewed the episode.

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/schematic}
\caption{\small \textbf{Topic weights in episode and recall content.} We used hand-annotated descriptions of each manually identified scene from the episode to fit a topic model.  Three example video frames (first row) and their associated descriptions (second row) are displayed.  The third row shows an example participant's later recalls of the same three scenes.  We used the topic model (fit to the episode annotations) to estimate topic vectors for each moment of the episode and each sentence of participants' recalls.  Example topic vectors are displayed in the bottom row (blue: episode annotations; green: example participant's recalls).  Three topic dimensions are shown (the highest-weighted topics for each of the three example scenes, respectively), along with the 10 highest-weighted words for each topic.  Figure~\topics~provides a full list of the top 10 words from each of the discovered topics.}
\label{fig:schematic}
\end{figure}
% add an x-label under the "

The 32 topics we found were heavily character-focused (i.e., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes centered around Sherlock Holmes (the titular character), John Watson (Sherlock's close confidant and assistant), supporting characters (e.g., Inspector Lestrade, Sergeant Donovan, or Sherlock's brother Mycroft), or the interactions between various groupings of these characters (see Fig.~\topics).  Several of the identified topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences if the distinctions between those overlapping topics were meaningful.  The topic vectors for each timepoint were also \textit{sparse}, in that only a small number (typically one or two) of topics tended to be ``active" in any given timepoint (see Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit \textit{persistence} (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with \textit{occasional rapid changes} (i.e., occasionally topics would appear to spring into or out of existence).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B) and reflect the gradual drift and sudden shifts fundamental to the temporal dynamics of real-world experiences.  Given this observation, we adapted an approach devised by \cite{BaldEtal17}, and used a hidden Markov model (HMM) to identify the \textit{event boundaries} where the topic activations changed rapidly (i.e., the boundaries of the blocks in the temporal correlation matrix; event boundaries identified by the HMM are outlined in yellow in Fig.~\ref{fig:model}B).  Part of our model fitting procedure required selecting an appropriate number of ``events" into which the topic trajectory should be segmented.  To accomplish this, we used an optimization procedure that maximized the difference between the topic weights for timepoints within an event versus timepoints across multiple events (see \textit{Methods} for additional details).  We then created a stable ``summary'' of the content within each episode event by averaging the topic vectors across the timepoints spanned by each event (Fig.~\ref{fig:model}C).

@@

Given that the time-varying content of the episode could be segmented cleanly into discrete events, we wondered whether participants' recalls of the episode also displayed a similar structure.  We applied the same topic model (already trained on the episode annotations) to each participant's recalls.  Analogously to how we parsed the time-varying content of the episode, to obtain similar estimates for each participant's recall, we treated each overlapping  window of (up to 10) sentences from their transcript as a ``document," and computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-windows by number-of-topics topic proportions matrix that characterized how the topics identified in the original episode were reflected in the participant's recalls.  Note that an important feature of our approach is that it allows us to compare participants' recalls to events from the original episode, despite different participants using widely varying language to describe the events, and that those descriptions often diverged in content and quality from the episode annotations.  This is a substantial benefit of projecting the episode and recalls into a shared ``topic'' space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

Although the example participant's recall topic proportions matrix has some visual similarity to the episode topic proportions matrix, the time-varying topic proportions for the example participant's recalls are not as sparse as those for the episode (compare Figs.~\ref{fig:model}A and D).  Similarly, although there do appear to be periods of stability in the recall topic dynamics (i.e., most topics are active or inactive over contiguous blocks of time), the changes in topic activations that define event boundaries appear less clearly delineated in participants' recalls than in the episode's annotations.  To examine these patterns in detail, we computed the timepoint-by-timepoint correlation matrix for the example participant's recall trajectory (Fig.~\ref{fig:model}E).  As in the episode correlation matrix (Fig.~\ref{fig:model}B), the example participant's recall correlation matrix has a strong block diagonal structure, indicating that their recalls are discretized into separated events.  As for the episode correlation matrix, we leveraged an HMM-based optimization procedure (see \textit{Methods}) to determine how many events are reflected in the participant's recalls and where specifically the event boundaries fall (outlined in yellow).  We carried out a similar analysis on all 17 participants' recall topic proportions matrices (Fig.~\corrmats).

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Fig.~\corrmats), each participant appeared to have a unique \textit{recall resolution}, reflected in the sizes of those blocks.  While some participants' recall topic proportions segmented into just a few events (e.g., Participants P4, P5, and P7), others' segmented into many shorter duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the episode with different levels of detail---i.e., some might touch on just the major plot points, whereas others might attempt to recall every minor scene or action.  The second clear pattern present in every individual participant's recall correlation matrix was that, unlike in the episode correlation matrix, there were substantial off-diagonal correlations.  Whereas each event in the original episode was (largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory, participants appeared to be integrating across multiple events, blending elements of previously recalled and not-yet-recalled content into each newly recalled event~\citep[Figs.~\ref{fig:model}E, \corrmats; also see][]{MannEtal11, HowaEtal12, Mann19}.

The above results indicate that both the structure of the original episode and participants' recalls of the episode exhibit event boundaries that can be identified automatically by characterizing the dynamic content using a shared topic model and segmenting the content into events via HMMs.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced in the episode, and the events they later recalled.  One approach to linking the experienced (episode) and recalled events is to label each recalled event as matching the episode event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yields a sequence of ``presented" events from the original episode, and a (potentially differently ordered) sequence of ``recalled" events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\ref{fig:list-learning}A;][]{AtkiShif68, PostPhil65, WelcBurn24}; how participants most often transition between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\ref{fig:list-learning}B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\ref{fig:list-learning}C;][]{Murd62a}. Interestingly, for two of these analyses (probability of first recall and lag-conditional response probability curves) we observed patterns comparable to classic effects from list-learning literature: namely, a higher probability of initiating recall with the first event in the sequence (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}B). In contrast, we did not observe a pattern comparable to the serial position effect (Fig.~\ref{fig:list-learning}C), but rather greater memory for specific events distributed approximately evenly throughout the episode.

We can also apply two list-learning-native analyses that describe how participants group items in their recall sequences: temporal clustering and semantic clustering \citep[][see \textit{Methods} for details]{PolyEtal09}.~Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed episode events were clustered heavily in participants' recall event sequences (mean clustering score: 0.767, SEM: 0.029), and that participants with higher temporal clustering scores tended to perform better according to both \cite{ChenEtal17}'s hand-annotated memory scores (Pearson's $r(15) = 0.62,~p = 0.008$) and our model's estimate (Pearson's $r(15) = 0.54,~p = 0.024$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  We found that participants tended to recall semantically similar episode events together (mean clustering score: 0.787, SEM: 0.018), and that semantic clustering score was also related to both hand-annotated  (Pearson's $r(15) = 0.65,~p = 0.004$) and model-derived (Pearson's $r(15) = 0.63,~p = 0.007$) memory performance.
% NOTE:
% Semantic and temporal clustering are both positively correlated with n_events -- r=0.49; p=0.046 and r=0.61; p=0.009, respectively.
% - Paxton

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} The episode-recall correlation matrix for a representative participant (17).  The yellow boxes highlight the maximum correlation in each column.  The example participant's overall precision score was computed as the average across correlation values in the yellow boxes.  Their distinctiveness score was computed as the average (over recall events) of 1 minus the average correlation between each recall event and all other recall events that do not display a box in the same row.  \textbf{B.} The (Pearson's) correlation between precision and hand-annotated memory performance. \textbf{C.} The correlation between distinctiveness and hand-annotated memory performance. \textbf{D.} The correlation between precision and the number of episode events successfully recalled, as determined by our model. \textbf{E.} The correlation between distinctiveness and the number of episode events successfully recalled, as determined by our model.}
  \label{fig:precision-distinctiveness}
\end{figure}
% Need to reconcile the description of distinctiveness with how we computed it... - Andy

Statistical models of memory studies often treat recall success as binary (in other words, an item either was or was not recalled), or occasionally categorical \citep[e.g., to distinguish familiarity from recollection;][]{YoneEtal02}.  Such approaches are tenable in classical list-learning or recognition memory paradigms, as the presented stimuli tend to be very simple (e.g., a sequence of individual words or items).  However, memory for naturalistic experiences is much more nuanced.  For example, certain aspects of an experience might be correctly remembered at varying levels of detail, or distorted, or forgotten entirely.  Further, each remembering is itself a richly structured phenomenon.  Our framework produces a content-based model of individual episode and recall events by projecting the dynamic content of the episode and participants' recalls into a shared topic space.  This allows for direct, quantitative comparisons between all stimulus and recall events, as well as between the recall events themselves.  Leveraging these content-based models of the stimulus/recall events, we developed two novel, \textit{continuous} metrics for analyzing naturalistic memory:~\textit{precision} and \textit{distinctiveness}.  Precision is intended to capture the ``completeness'' of recall, or how fully the presented content was recapitulated in memory.  We define a recall event's precision as the maximum correlation between the topic proportions of that recall event and any episode event (Fig.~\ref{fig:precision-distinctiveness}).  A second novel metric we introduce here is \textit{distinctiveness}, which is intended to capture the ``specificity'' of recall.  In other words, distinctiveness quantifies the extent to which a given recalled event reflects the most similar presented event moreso than it does other presented events.  To compute a recall event's distinctiveness, we first identify the episode event to which its topic vector is most strongly correlated.  We then define distinctiveness as one minus the average correlation between the given recall event and all \textit{other} episode events.  In addition to individual events, one may also use these metrics to describe each participant's overall performance by averaging across a participant's event-wise precision or distinctiveness scores.  Participants whose recall events are more veridical descriptions of what happened in the episode event will presumably have higher precision scores. We find that, across participants, higher precision scores are positively correlated with both hand-annotated memory performance (as collected by Chen et al., 2017; Pearson's $r(15) = 0.60, p = 0.010$) and the number of episode events successfully remembered, as determined by our model (Pearson's $r(15) = 0.90, p < 0.001$).  We also hypothesized that participants who recounted events in a more distinctive way would display better overall memory.  We find that participants' distinctiveness scores were positively
correlated with our model's estimated number of recall events (Pearson's $r(15) = 0.55, p = 0.021$).  However, we found no evidence that distinctiveness scores were correlated with hand-annotated memory performance (Pearson's $r(15) = 0.28, p = 0.275$).  We elaborate on this potential discrepency in the \textit{Discussion} section.

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_detail}
  \caption{\small \textbf{Precision metric reflects completeness of recall.} \textbf{A.} Recall precision by episode event.  Grey violin plots display kernel density estimates for the distribution of recall precision scores for a single episode event.  Colored dots within each violin plot represent individual participants' recall precision for the given event.  episode events are ordered along the $x$-axis by the average precision with which they were remembered.  \textbf{B.} The set of ``Narrative Details" episode annotations (generated by \citealp{ChenEtal17}) for scenes comprising an example episode event (22) identified by the HMM.  Each action or feature is highlighted in a different color.  \textbf{C.} A subset of the sentences comprising the most precise (P17) and least precise (P6) participants' recalls of episode event 22.  Descriptions of specific actions or features reflecting those highlighted in panel B are highlighted in the corresponding color.}
  \label{fig:precision-detail}
\end{figure}

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/distinctiveness_detail}
  \caption{\small \textbf{Distinctiveness metric reflects specificity of recall.} \textbf{A.} Recall distinctiveness by episode event.  Kernel density estimates for each episode event's distribution of recall distinctiveness scores, analogous to Fig.~\ref{fig:precision-detail}A.  \textbf{B.} The sets of ``Narrative Details" episode annotations (generated by \citealp{ChenEtal17}) for scenes comprising episode events described by the example participants in panel C.  Each event's text is highlighted in a different color.  \textbf{C.} The sentences comprising the most distinctive (P13) and least distinctive (P11) participants' recalls of episode event 19.  Sections of recall describing each each episode event in panel B are highlighted with the corresponding color.}
  \label{fig:distinctiveness-detail}
\end{figure}

Further intuition for the behaviors captured by these two metrics may be gained by directly examining the content of the episode and recalls our framework models.  In Figure \ref{fig:precision-detail}, we contrast recalls for the same episode event (event 22) from two participants: one with a high precision score (P17), the other with a low precision score (P6).  From the HMM-identified event boundaries, we recovered the set of annotations describing the content of an example episode event (Fig.~\ref{fig:precision-detail}B), and divided them into different color-coded sections for each action or feature described.  We then similarly recovered the set of sentences comprising the corresponding recall event for each of the two example participants.  Because the recall sliding windows overlap heavily, and each recall event spans multiple recall timepoints (i.e., windows), we have stripped any sentences from the beginning and end that describe earlier or later episode events for the sake of readability.  In other words, Fig.~\ref{fig:precision-detail}C shows a subset of the full recall event text, comprising sentences between the first and last descriptions of content from the example episode event.  We then colored all words describing actions and features coded in panel B by their corresponding color.  Visual comparison of these example transcripts reveals that the more precise recall captures more of the episode event's content, and with more detail.

>>>>

Occasionally, this gradual ``drift'' of our ongoing experience is punctuated by sudden changes, or ``shifts''~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) help to discretize our experiences (and their mental representations) into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within-event) and transient (across-event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distills those experiences down to their essence.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.  This work also suggests a means of distinguishing the essence of an experience from its low-level details.  The overall structure of events and event transitions reflects how the high-level experience unfolds (i.e., its essence), while subtler event-level properties reflect low-level details.  Prior research has also implicated a network of brain regions (including the hippocampus and the medial prefrontal cortex) in playing a critical role in transforming experiences into structured and consolidated memories ~\citep{TompDava17}.

@@

\cite{ChenEtal17} collected hand-annotated information about each of 1,000 (manually identified) scenes spanning the roughly 50 minute video used in their experiment.  This information included: a brief narrative description of what was happening, the location where the scene took place, the names of any characters on the screen, and other similar details (for a full list of annotated features, see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features and scenes as the ``vocabulary'' for the topic model.  We then concatenated the sets of words across all features contained in overlapping sliding windows of (up to) 50 scenes, and treated each window as a single ``document'' for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the episode (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  Note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06}, in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each sliding window of annotations transformed by the topic model.  We then stretched (interpolated) the resulting windows-by-topics matrix to match the time series of the 1,976 fMRI volumes collected as participants viewed the episode.

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/schematic}
\caption{\small \textbf{Topic weights in episode and recall content.} We used hand-annotated descriptions of each manually identified scene from the episode to fit a topic model.  Three example episode frames (first row) and their associated descriptions (second row) are displayed.  The third row shows an example participant's later recalls of the same three scenes.  We used the topic model (fit to the episode annotations) to estimate topic vectors for each moment of the episode and each sentence of participants' recalls.  Example topic vectors are displayed in the bottom row (blue: episode annotations; green: example participant's recalls).  Three topic dimensions are shown (the highest-weighted topics for each of the three example scenes, respectively), along with the 10 highest-weighted words for each topic.  Figure~\topics~provides a full list of the top 10 words from each of the discovered topics.}
\label{fig:schematic}
\end{figure}
% add an x-label under the "

The 32 topics we found were heavily character-focused (i.e., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes centered around Sherlock Holmes (the titular character), John Watson (Sherlock's close confidant and assistant), supporting characters (e.g., Inspector Lestrade, Sergeant Donovan, or Sherlock's brother Mycroft), or the interactions between various groupings of these characters (see Fig.~\topics).  This likely follows from the frequency with which these terms appeared in the episode annotations.  Several of the identified topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences if the distinctions between those overlapping topics were meaningful.  The topic vectors for each timepoint were also \textit{sparse}, in that only a small number (typically one or two) of topics tended to be ``active'' in any given timepoint (see Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit \textit{persistence} (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with \textit{occasional rapid changes} (i.e., occasionally topics would appear to spring into or out of existence).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B) and reflect the gradual drift and sudden shifts fundamental to the temporal dynamics of many real-world experiences, as well as television episodes.  Given this observation, we adapted an approach devised by \cite{BaldEtal17}, and used a hidden Markov model (HMM) to identify the \textit{event boundaries} where the topic activations changed rapidly (i.e., the boundaries of the blocks in the temporal correlation matrix; event boundaries identified by the HMM are outlined in yellow in Fig.~\ref{fig:model}B).  Part of our model fitting procedure required selecting an appropriate number of ``events'' into which the topic trajectory should be segmented.  To accomplish this, we used an optimization procedure that maximized the difference between the topic weights for timepoints within an event versus timepoints across multiple events (see \textit{Methods} for additional details).  We then created a stable ``summary'' of the content within each episode event by averaging the topic vectors across the timepoints spanned by each event (Fig.~\ref{fig:model}C).

@@

Given that the time-varying content of the episode could be segmented cleanly into discrete events, we wondered whether participants' recalls of the episode also displayed a similar structure.  We applied the same topic model (already trained on the episode annotations) to each participant's recalls.  Analogously to how we parsed the time-varying content of the episode, to obtain similar estimates for each participant's recall transcript, we treated each overlapping  window of (up to 10) sentences from their transcript as a ``document,'' and computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-windows by number-of-topics topic proportions matrix that characterized how the topics identified in the original episode were reflected in the participant's recalls.  An important feature of our approach is that it allows us to compare participants' recalls to events from the original episode, despite that different participants used widely varying language to describe the events, and that those descriptions often diverged in content and quality from the episode annotations.  This ability to match up conceptually related text that differs in specific vocabulary, detail, and length is an important benefit of projecting the episode and recalls into a shared ``topic'' space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

Although the example participant's recall topic proportions matrix has some visual similarity to the episode topic proportions matrix, the time-varying topic proportions for the example participant's recalls are not as sparse as those for the episode (compare Figs.~\ref{fig:model}A and D).  Similarly, although there do appear to be periods of stability in the recall topic dynamics (i.e., most topics are active or inactive over contiguous blocks of time), the changes in topic activations that define event boundaries appear less clearly delineated in participants' recalls than in the episode's annotations.  To examine these patterns in detail, we computed the timepoint-by-timepoint correlation matrix for the example participant's recall trajectory (Fig.~\ref{fig:model}E).  As in the episode correlation matrix (Fig.~\ref{fig:model}B), the example participant's recall correlation matrix has a strong block diagonal structure, indicating that their recalls are discretized into separated events.  We used the same HMM-based optimization procedure that we had applied to the episode's topic proportions matrix (see \textit{Methods}) to estimate an analogous set of event boundaries in the participant's recounting of the episode (outlined in yellow).  We carried out this analysis on all 17 participants' recall topic proportions matrices (Fig.~\corrmats).

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Fig.~\corrmats), each participant appeared to have a unique \textit{recall resolution}, reflected in the sizes of those blocks.  While some participants' recall topic proportions segmented into just a few events (e.g., Participants P4, P5, and P7), others' segmented into many shorter duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the episode with different levels of detail---i.e., some might recount only high-level essential plot details, whereas others might recount low-level details instead (or in addition).  The second clear pattern present in every individual participant's recall correlation matrix was that, unlike in the episode correlation matrix, there were substantial off-diagonal correlations.  Whereas each event in the original episode was (largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory, participants appeared to be integrating across multiple events, blending elements of previously recalled and not-yet-recalled content into each newly recalled event~\citep[Figs.~\ref{fig:model}E, \corrmats; also see][]{MannEtal11, HowaEtal12, Mann19}.

The above results demonstrate that topic models capture the dynamic conceptual content of the episode and participants' recalls of the episode.  Further, the episode and recalls exhibit event boundaries that can be identified automatically using HMMs to segment the dynamic content.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced in the episode, and the events they later recalled.  We labeled each recalled event as matching the episode event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yielded a sequence of ``presented'' events from the original episode, and a (potentially differently ordered) sequence of ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\ref{fig:list-learning}A;][]{AtkiShif68, PostPhil65, WelcBurn24}; how participants most often transition between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\ref{fig:list-learning}B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\ref{fig:list-learning}C;][]{Murd62a}. Some of the patterns we observed appeared to be similar to classic effects from the list-learning literature.  For example, participants had a higher probability of initiating recall with the first event in the sequence (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}B). However, unlike what is typically observed in list-learning studies, we did not observe patterns comparable to the primacy or recency serial position effects (Fig.~\ref{fig:list-learning}C).  We hypothesized that participants might be leveraging the meaningful narrative associations and references over long timescales throughout the episode.

Clustering scores are often used by memory researchers to characterize how people organize their memories of words on a studied list~\citep[for review, see][]{PolyEtal09}.  We defined analogous measures to characterize how participants organized their memories for episodic events (see \textit{Methods} for details).  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed episode events tended to appear nearby in participants' recall event sequences (mean clustering score: 0.767, SEM: 0.029).  Participants with higher temporal clustering scores tended to exhibit better overall memory for the episode, according to both \cite{ChenEtal17}'s hand-counted numbers of recalled scenes from the episode (Pearson's $r(15) = 0.62,~p = 0.008$) and the numbers of episode events that best-matched at least one recalled event (i.e., model-estimated number of recalled events; Pearson's $r(15) = 0.49, p = 0.0046$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  We found that participants tended to recall semantically similar episode events together (mean clustering score: 0.787, SEM: 0.018), and that semantic clustering score was also related to both hand-annotated  (Pearson's $r(15) = 0.65,~p = 0.004$) and model-estimated (Pearson's $r(15) = 0.61,~p = 0.0092$) numbers of recalled events.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} The episode-recall correlation matrix for a representative participant (17).  The yellow boxes highlight the maximum correlation in each column.  The example participant's overall precision score was computed as the average across correlation values in the yellow boxes.  Their distinctiveness score was computed as the average (over recall events) of 1 minus the average correlation between each recall event and all other recall events that do not display a box in the same row.  \textbf{B.} The (Pearson's) correlation between precision and hand-annotated memory performance. \textbf{C.} The correlation between distinctiveness and hand-annotated memory performance. \textbf{D.} The correlation between precision and the number of episode events successfully recalled, as determined by our model. \textbf{E.} The correlation between distinctiveness and the number of episode events successfully recalled, as determined by our model.}
  \label{fig:precision-distinctiveness}
\end{figure}

The above analyses illustrate how our framework for characterizing the dynamic conceptual content of naturalistic episodes enables us to carry out analyses that have traditionally been applied to much simpler list-learning paradigms.  However, perhaps the most interesting aspects of memory for naturalistic episodes are those that have no list-learning analogs.  The nuances of how one's memory for an event might capture some details, yet distort or neglect others, is central to how we use our memory systems in daily life.  Yet when researchers study memory in highly simplified paradigms, those nuances are not typically observable.  We next developed two novel continuous metrics, termed precision and distinctiveness, aimed at characterizing distortions in the conceptual content of individual recalled events, and the conceptual overlap between how people described different events.

\textit{Precision} is intended to capture the ``completeness'' of recall, or how fully the presented content was recapitulated in memory.  We define a recall event's precision as the maximum correlation between the topic proportions of that recall event and any episode event (Fig.~\ref{fig:precision-distinctiveness}).  In other words, given that a recalled event best matches a particular episode event, more precisely recalled events overlap more strongly with the conceptual content of the original episode event.  When a given event is assigned a blend of several topics, as is often the case (Fig.~\ref{fig:model}), a high precision score requires recapitulating the relative topic proportions during recall.

A second novel metric we developed is \textit{distinctiveness}, which is intended to capture the ``specificity'' of recall.  In other words, distinctiveness quantifies the extent to which a given recalled event reflects the most similar episode event over and above its reflection of other episode events.  Intuitively, distinctiveness is like a normalized variant of our precision metric.  Whereas precision solely measures how much detail about an episode was captured in someone's recall, distinctiveness penalizes details that also pertain to other episode events.  We define the distinctiveness of an event's recall as its precision expressed in standard deviation units with respect to other episode events.  Specifically, for a given recall event, we compute the correlation between its topic vector and that of each episode event.  This yields a distribution of correlation coefficients (one per episode event).  We subtract the mean and divide by the standard deviation of this distribution to $z$-score the coefficients.  The maximum value in this distribution (which, by definition, belongs to the episode event that best matches the given recall event) is the recall event's distinctiveness score.  In this way, recall events that match one episode event far better than all other episode events will receive a high distinctiveness score.  By contrast, a recall event that matches all episode events roughly equally will receive a comparatively low distinctiveness score.

In addition to examining how precisely and distinctively participants recalled individual events, one may also use these metrics to summarize each participant's performance by averaging across a participant's event-wise precision or distinctiveness scores.  This enables us to quantify how precisely a participant tended to recall subtle within-event details, as well as how specific (distinctive) those details were to individual events from the episode.  Participants' average precision and distinctiveness scores were strongly correlated ($r(15) = 0.90, p < 10^{-5}$).  This indicates that participants who tended to precisely recount low-level details of episode events also tended to do so in an event-specific way (e.g., as opposed to detailing recurring themes that were present in most or all episode events; this strategy would have resulted in high precision but low distinctiveness).  We found that, across participants, higher precision scores were positively correlated with both the hand-annotated ($r(15) = 0.60, p = 0.010$) and model-estimated ($r(15) = 0.90, p < 0.001$) numbers of events that participants recalled.  Participants' average distinctiveness scores were also correlated with both the hand-annotated ($r(15) = 0.45, p = 0.068$) and model-estimated ($r(15) = 0.71, p = 0.001$) numbers of recalled events.

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_detail}
  \caption{\small \textbf{Precision metric reflects completeness of recall.} \textbf{A.} Recall precision by episode event.  Grey violin plots display kernel density estimates for the distribution of recall precision scores for a single episode event.  Colored dots within each violin plot represent individual participants' recall precision for the given event.  episode events are ordered along the $x$-axis by the average precision with which they were remembered.  \textbf{B.} The set of ``Narrative Details'' episode annotations (generated by \citealp{ChenEtal17}) for scenes comprising an example episode event (22) identified by the HMM.  Each action or feature is highlighted in a different color.  \textbf{C.} A subset of the sentences comprising the most precise (P17) and least precise (P6) participants' recalls of episode event 22.  Descriptions of specific actions or features reflecting those highlighted in panel B are highlighted in the corresponding color.}
  \label{fig:precision-detail}
\end{figure}

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/distinctiveness_detail}
  \caption{\small \textbf{Distinctiveness metric reflects specificity of recall.} \textbf{A.} Recall distinctiveness by episode event.  Kernel density estimates for each episode event's distribution of recall distinctiveness scores, analogous to Fig.~\ref{fig:precision-detail}A.  \textbf{B.} The sets of ``Narrative Details'' episode annotations (generated by \citealp{ChenEtal17}) for scenes comprising episode events described by the example participants in panel C.  Each event's text is highlighted in a different color.  \textbf{C.} The sentences comprising the most distinctive (P13) and least distinctive (P11) participants' recalls of episode event 19.  Sections of recall describing each each episode event in panel B are highlighted with the corresponding color.}
  \label{fig:distinctiveness-detail}
\end{figure}
